Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022

### 2025-06-25 manuscript/manuscript.tex
Working on word doc

%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
%%\documentclass[sn-mathphys,Numbered]{sn-jnl}% Math and Physical Sciences Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
\documentclass[lineno,sn-vancouver,Numbered,referee]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
%%\documentclass[default]{sn-jnl}% Default
%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{lstautogobble}%
\usepackage{makecell}%
%%%%

@@

VA involves two major components: survey and COD assignment \citep{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained interviewers use structured questionnaires and open narrative prompts to gather data from relatives or close contacts of the deceased. In the COD assignment component, physicians review these data to determine the most likely COD. However, reliance on physician assignment has been criticized for limited reproducibility and subjectivity \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005,mccormickProbabilisticCauseofDeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. To overcome these limitations, automated Computer Coded Verbal Autopsy (CCVA) methods such as InterVA \citep{byassIntegratedApproachProcessing2019} and InSilicoVA \citep{mccormickProbabilisticCauseofDeathAssignment2016b} have been developed. These models offer scalable and reproducible alternatives and have demonstrated comparable performance to physicians at the population level. However, their performance at the individual level remains limited \citep{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}, while their reliance on structured questionnaire data often omits open narrative text, which can contain additional contextual and chronological information that may improve diagnostic accuracy \citep{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

@@

This section outlines key limitations of the current study related to the use of GPT models. Section \ref{discuss-limits-phy} discusses the use of physician assignment as the reference standard for comparing models. Section \ref{discuss-limits-model} addresses the need for further parameter tuning and the evaluation of model consistency and multi-COD assignments, while Section \ref{discuss-limits-global} notes the importance of more diverse datasets for model evaluation.

@@

\subsubsection{Model Tuning, Consistency, and Multiple Outputs}\label{discuss-limits-model}

GPT-3.5/4 were used with default parameters except for temperature, which was set to 0 to enhance consistency. However, tuning temperature and other settings could potentially improve performance \citep{openaiOpenAIPlatformAPI2024}, but was not explored due to the high cost of repeated runs needed for sensitivity analyses, as noted in Section \ref{discuss-disadv-reprod}. Despite temperature control, GPT outputs may still vary, highlighting the need to assess reliability and consistency to avoid coincidental results \citep{johnsonAssessingAccuracyReliability2023,jangConsistencyAnalysisChatGPT2023,krishnaEvaluationReliabilityRepeatability2024}. Unlike GPT models, InterVA-5 and InSilicoVA provide multiple COD assignments with associated probabilities to measure reliability. In addition, while GPT can be prompted to generate multiple CODs, this study evaluated only the most probable assignment. Considering multiple COD outputs may better capture alternative diagnoses and align more closely with physician assessments \citep{solemanVerbalAutopsyCurrent2006}.

>>>>

%%\documentclass[sn-nature]{sn-jnl}% Style for submissions to Nature Portfolio journals
%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
%%\documentclass[sn-mathphys,Numbered]{sn-jnl}% Math and Physical Sciences Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
\documentclass[sn-vancouver,Numbered,referee]{sn-jnl}% Vancouver Reference Style
%%\documentclass[sn-apa]{sn-jnl}% APA Reference Style 
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
%%\documentclass[default]{sn-jnl}% Default
%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout

%%%% Standard Packages
%%<additional latex packages if required can be included here>
\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{lstautogobble}%
\usepackage{makecell}%
%%%%

@@

VA involves two major components: survey and COD assignment \citep{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained interviewers use structured questionnaires and open narrative prompts to gather data from relatives or close contacts of the deceased. In the COD assignment component, physicians review these data to determine the most likely COD. However, reliance on physician assignment has been criticized for limited reproducibility and subjectivity \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005,mccormickProbabilisticCauseofdeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. To overcome these limitations, automated Computer Coded Verbal Autopsy (CCVA) methods such as InterVA \citep{byassIntegratedApproachProcessing2019} and InSilicoVA \citep{mccormickProbabilisticCauseofDeathAssignment2016b} have been developed. These models offer scalable and reproducible alternatives and have demonstrated comparable performance to physicians at the population level. However, their performance at the individual level remains limited \citep{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}, while their reliance on structured questionnaire data often omits open narrative text, which can contain additional contextual and chronological information that may improve diagnostic accuracy \citep{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

@@

This section outlines key limitations of the current study related to the use of GPT models. Section \ref{discuss-limits-phy} discusses the use of physician assignment as the reference standard for comparing models. Section \ref{discuss-limits-model} addresses the need for further model tuning and adjustments of model parameters and output, while Section \ref{discuss-limits-global} notes the importance of more globally diverse datasets for model evaluation.

@@

\subsubsection{Model Tuning and Interpretability}\label{discuss-limits-model}

GPT-3.5/4 were used with default parameters except for temperature, which was set to 0 to enhance consistency. However, tuning temperature and other settings could potentially improve performance \citep{openaiOpenAIPlatformAPI2024}, but was not explored due to the high cost of repeated runs needed for sensitivity analyses, as noted in Section \ref{discuss-disadv-reprod}. Despite temperature control, GPT outputs may still vary, highlighting the need to assess reliability and consistency to avoid coincidental results \citep{johnsonAssessingAccuracyReliability2023,jangConsistencyAnalysisChatGPT2023,krishnaEvaluationReliabilityRepeatability2024}. Unlike GPT models, InterVA-5 and InSilicoVA provide multiple COD assignments with associated probabilities to measure reliability. In addition, while GPT can be prompted to generate multiple CODs, this study evaluated only the most probable assignment. Considering multiple COD outputs may better capture alternative diagnoses and align more closely with physician assessments \citep{solemanVerbalAutopsyCurrent2006}. Notably, GPT models are capable of generating natural language justifications for their COD assignments, offering a form of qualitative interpretability that mimics physician reasoning. However, these explanations are not tied to explicit probabilistic scores or model certainty, which limits their use for structured reliability assessments in current VA workflows. Nonetheless, evaluation methods considering probability scores and justifications may identifying insights on model performance and behavior differences across CODs.
